---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{Getting Started}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse  = TRUE,
    comment   = "#>",
    fig.width = 7.5
)
```

Start by loading `pwcuratr` and a collection of packages that 
facilitates data manipulation and plotting.

```{r setup}
library(pwcuratr)
library(glue)
library(tidyverse)
library(pander)
```

The first step to curating a pathway cluster is to define a 
set of 'seed genes' that are of interest.
This set could, e.g., be defined via an extensive literature review.
The package `pwcuratr` comes with a set of genes associated with
serotonin neurotransmitter pathways.
The list of genes can be loaded via

```{r}
seed_genes <- read_csv(
    glue("{find.package('pwcuratr')}/data/seed_genes.csv"),
    col_types = c(ensembl_gene_id = col_character())
)$ensembl_gene_id

print(seed_genes)
```

Next, external data is used to define a plausible functional
neighborhood for genes that are functionally related to 
any of the seed genes.
We start by querying a snapshot of the reactome.org pathway
database for any pathways containing at least one of the seed genes.

```{r}
tbl_reactome_pathways <- tibble(
    reactome_pathway_id = query_reactome_pathways(seed_genes),
    n_participating_genes = map_int(
        reactome_pathway_id,
        ~length(query_participating_genes(.))
    )
) %>% 
arrange(n_participating_genes)

pander(tbl_reactome_pathways)
```

A naive approach to defining a function neighborhood for the seed
genes would be to take the union of all genes who are participating
in any of the identified reactome pathways.
However, reactome.org pathways are defined on very different 
levels of granularity and it will not always be sensible to include
all participating genes from very large pathways.
A first step to address this 'specificity problem' is to hand-curate
the list of pathways to consider for further processing.
Here, a completely arbitrary cut-off in terms of the pathway 
size of 250 participating genes is chosen.
To avoid loosing any of the seed genes, they are added manually to
the initial list of candidate genes.

```{r}
candidate_genes <- c(
    query_participating_genes(
        tbl_reactome_pathways %>% 
            filter(n_participating_genes <= 250) %>% 
            pull(reactome_pathway_id)
    ),
    seed_genes
) %>% 
unique


length(candidate_genes)
```

To refine the candidate set of genes even further, 
predicted functional gene-gene interactions based on the 2018
release of the interactions derived in

> Wu, G., Feng, X., & Stein, L. (2010). A human functional protein interaction network and its application to cancer data analysis. Genome biology, 11(5), R53.

are used.
The set of interactions is scored between 0 (low confidence) and 1 
(high confidence) for functional interaction between the 
respective genes.
```{r}
pwcuratr_tbls$reactome_interactions %>% 
    ggplot() + 
    aes(score) + 
    geom_histogram(binwidth = .01) +
    ggtitle("Confidence score distribution for functional interactions") +
    theme_bw()
```

Clearly, the majority of included interactions has a confidence
score of 0.9 or greater and we filter the set of interactions 
with this threshold.
This results in a gene interaciton network that can be used to
refine our notion of funcional neighborhood.
We want to filter the set of candidate genes derived above such 
that only genes that are connected to a seed gene via at most $k$
edges are retained.

```{r}
tibble(
    maxedgedistance = 0:10,
    n_genes_pruned  = map_int(
        maxedgedistance,
        ~length(prune(
            candidate_genes, 
            seed_genes, 
            minscore        = .9, 
            maxedgedistance = .
        ))
    )
) %>% 
ggplot() +
    aes(maxedgedistance, n_genes_pruned) +
    geom_line() + 
    geom_point() +
    scale_y_continuous(
        "number of retained genes",
        limits = c(0L, NA_integer_)
    ) + 
    scale_x_continuous("maximal edge distance", breaks = 0:10) +
    ggtitle("Number of retained candiate genes by choice of maximal edge-distance (k)") +
    theme_bw() + 
    theme(
        panel.grid.minor = element_blank()
    )
```

The final choice of $k$ can only be justified by manually inspecting 
the resulting gene-networks.
As an example, consider the choice $k=2$.
We first prune the set of candidate genes and then create an
igraph object to represent the network information.
This allows us to use the wealth of graph-based techniques
implemented in `igraph` package.
For instance, one may be interested in the number of connected 
components of the pruned gene network.

```{r}
candidate_genes <- prune(
    candidate_genes, 
    seed_genes, 
    minscore        = .9, 
    maxedgedistance = 2
) 

gr <- as_igraph(candidate_genes, minscore = .9)

igraph::components(gr)$csize
```

There are 7 components with almost all candidate genes connected in 
the first components and two small non-trivial components.
All components of size 1 correspond to seed genes for which no 
functional neighbors in the selected reactome.org pathways 
where found. 

```{r, fig.width=12, fig.height=3*3*6}
plot_pathway(
    candidate_genes, 
    seed_genes, 
    minscore  = 0.9, 
    titlesize = 20
)
```

Based on this visual representation, the neighborhood parameters could 
be tweaked or the set of initial reactome pathways could be altered.
In the presence of non-connected components it might be worthwhile 
considering a split of the pathway cluster such that each individual
network is fully connected.

For instance, one could consider the HTR subsystem separately by 
restricting the initial set of seed genes.
Since the HTR3 subsystem is not connected, we can filter these seed 
genes out as well.

```{r}
seed_genes_htr <- pwcuratr_tbls$ensemble %>% 
    select(external_gene_name, ensembl_gene_id) %>% 
    distinct() %>% 
    filter(
        ensembl_gene_id %in% seed_genes, 
        str_detect(external_gene_name, "HTR[^3]")
    ) %>% 
    pull(ensembl_gene_id) %>% 
    unique()

tbl_reactome_pathways_htr <- tibble(
        reactome_pathway_id = query_reactome_pathways(seed_genes_htr),
        n_participating_genes = map_int(
            reactome_pathway_id,
            ~length(query_participating_genes(.))
        )
    ) %>% 
    arrange(n_participating_genes)

candidate_genes_htr <- c(
        query_participating_genes(
            tbl_reactome_pathways_htr %>% 
                filter(n_participating_genes <= 250) %>% 
                pull(reactome_pathway_id)
        ),
        seed_genes_htr
    ) %>% 
    unique %>% 
    prune(
        seed_genes_htr, 
        minscore        = .9, 
        maxedgedistance = 2
    ) 

gr <- as_igraph(candidate_genes_htr, minscore = .9)

igraph::components(gr)$csize
```

```{r, fig.width=12, fig.height=12}
plot_graph(
    gr, 
    seed_genes,
    titlesize = 20
)
```
